---
title: "TIFN2: Red fluorescent plaque development in gingivitis"
output: html_document
date: "2025-08-31"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preamble

```{r setup}
library(dplyr)
library(tidyr)
library(parafac4microbiome)
library(NPLStoolbox)
library(CMTFtoolbox)
library(ggplot2)
library(ggpubr)
library(scales)
library(ggpattern)
```

```{r Import red fluorescence data}
rf_data = read.csv("./TIFN2/RFdata.csv")
colnames(rf_data) = c("subject", "id", "fotonr", "day", "group", "RFgroup", "MQH", "SPS(tm)", "Area_delta_R30", "Area_delta_Rmax", "Area_delta_R30_x_Rmax", "gingiva_mean_R_over_G", "gingiva_mean_R_over_G_upper_jaw", "gingiva_mean_R_over_G_lower_jaw")
rf_data = rf_data %>% as_tibble()

rf_data[rf_data$subject == "VSTPHZ", 1] = "VSTPH2"
rf_data[rf_data$subject == "D2VZH0", 1] = "DZVZH0"
rf_data[rf_data$subject == "DLODNN", 1] = "DLODDN"
rf_data[rf_data$subject == "O3VQFX", 1] = "O3VQFQ"
rf_data[rf_data$subject == "F80LGT", 1] = "F80LGF"
rf_data[rf_data$subject == "26QQR0", 1] = "26QQrO"

rf_data2 = read.csv("./TIFN2/red_fluorescence_data.csv") %>% as_tibble()
rf_data2 = rf_data2[,c(2,4,181:192)]
rf_data = rf_data %>% left_join(rf_data2)

rf = rf_data %>% select(subject, RFgroup) %>% unique()
```

```{r Import subject metadata}
age_gender = read.csv("./TIFN2/Ploeg_subjectMetadata.csv", sep=";")
age_gender = age_gender[2:nrow(age_gender),2:ncol(age_gender)]
age_gender = age_gender %>% as_tibble() %>% filter(onderzoeksgroep == 0) %>% select(naam, leeftijd, geslacht)
colnames(age_gender) = c("subject", "age", "gender")

# Correction for incorrect subject ids
age_gender[age_gender$subject == "VSTPHZ", 1] = "VSTPH2"
age_gender[age_gender$subject == "D2VZH0", 1] = "DZVZH0"
age_gender[age_gender$subject == "DLODNN", 1] = "DLODDN"
age_gender[age_gender$subject == "O3VQFX", 1] = "O3VQFQ"
age_gender[age_gender$subject == "F80LGT", 1] = "F80LGF"
age_gender[age_gender$subject == "26QQR0", 1] = "26QQrO"

age_gender = age_gender %>% arrange(subject)
```

```{r metadata testing}

mapping = c(-14,0,2,5,9,14,21)

testMetadata = function(model, comp, metadata){
  transformedSubjectLoadings = transformPARAFACloadings(model$Fac, 1)[,comp]
  transformedSubjectLoadings = transformedSubjectLoadings / norm(transformedSubjectLoadings, "2")
  
  metadata = metadata$mode1 %>% left_join(vanderPloeg2024$red_fluorescence %>% select(subject,day,Area_delta_R30,plaquepercent,bomppercent) %>% filter(day==14), by="subject")
  
  result = lm(transformedSubjectLoadings ~ plaquepercent + bomppercent + Area_delta_R30 + gender + age, data=metadata)
  
  # Extract coefficients and confidence intervals
  coef_estimates <- summary(result)$coefficients
  conf_intervals <- confint(result)
  
  # Remove intercept
  coef_estimates <- coef_estimates[rownames(coef_estimates) != "(Intercept)", ]
  conf_intervals <- conf_intervals[rownames(conf_intervals) != "(Intercept)", ]
  
  # Combine into a clean data frame
  summary_table <- data.frame(
    Term     = rownames(coef_estimates),
    Estimate = coef_estimates[, "Estimate"] * 1e3,
    CI       = paste0(
                 conf_intervals[, 1], " – ",
                 conf_intervals[, 2]
               ),
    P_value  = coef_estimates[, "Pr(>|t|)"],
    P_adjust = p.adjust(coef_estimates[, "Pr(>|t|)"], "BH"),
    row.names = NULL
  )
  
  return(summary_table)
}

testFeatures = function(model, metadata, componentNum, metadataVar){
  df = metadata$mode1 %>% left_join(rf_data %>% filter(day==14)) %>% left_join(age_gender)
  topIndices = metadata$mode2 %>% mutate(index=1:nrow(.), Comp = model$Fac[[2]][,componentNum]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
  bottomIndices = metadata$mode2 %>% mutate(index=1:nrow(.), Comp = model$Fac[[2]][,componentNum]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()
  
  timepoint = which(abs(model$Fac[[3]][,componentNum]) == max(abs(model$Fac[[3]][,componentNum])))
  
  Xhat = parafac4microbiome::reinflateTensor(model$Fac[[1]][,componentNum], model$Fac[[2]][,componentNum], model$Fac[[3]][,componentNum])
  y = df[metadataVar]
  
  print("Positive loadings:")
  print(cor(Xhat[,topIndices[1],timepoint], y))
  print("Negative loadings:")
  print(cor(Xhat[,bottomIndices[1],timepoint], y))
}

plotFeatures = function(mode2, model, componentNum, flip=FALSE){
  
  if(flip==TRUE){
    df = mode2 %>% mutate(Comp = -1*model$Fac[[2]][,componentNum]) %>% arrange(Comp) %>% filter(Species != "") %>% mutate(index=1:nrow(.), name = paste(Genus, Species))
  } else{
    df = mode2 %>% mutate(Comp = model$Fac[[2]][,componentNum]) %>% arrange(Comp) %>% filter(Species != "") %>% mutate(index=1:nrow(.), name = paste(Genus, Species))
  }
  df = rbind(df %>% head(n=10), df %>% tail(n=10))
  
  plot=df %>%
    ggplot(aes(x=Comp,y=as.factor(index),fill=as.factor(Phylum))) +
    geom_bar(stat="identity", col="black") +
    scale_y_discrete(label=df$name) +
    xlab("Loading") +
    ylab("") +
    guides(fill=guide_legend(title="Phylum")) +
    theme(text=element_text(size=16))
  
  return(plot)
}
```

```{r define colours}
phylum_colors = hue_pal()(7)
```

# ACMTF

## Process

```{r acmtf homogenize and process data}
homogenizedSubjects = parafac4microbiome::vanderPloeg2024$metabolomics$mode1$subject
mask = parafac4microbiome::vanderPloeg2024$tongue$mode1$subject %in% homogenizedSubjects

# Tongue
temp = parafac4microbiome::vanderPloeg2024$tongue
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedTongue = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Lowling
temp = parafac4microbiome::vanderPloeg2024$lower_jaw_lingual
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedLowling = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Lowinter
temp = parafac4microbiome::vanderPloeg2024$lower_jaw_interproximal
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedLowinter = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Upling
temp = parafac4microbiome::vanderPloeg2024$upper_jaw_lingual
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedUpling = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Upinter
temp = parafac4microbiome::vanderPloeg2024$upper_jaw_interproximal
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedUpinter = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Saliva
temp = parafac4microbiome::vanderPloeg2024$saliva
temp$data = temp$data[mask,,]
temp$mode1 = temp$mode1[mask,]
processedSaliva = processDataCube(temp, sparsityThreshold=0.50, considerGroups=TRUE, groupVariable="RFgroup", CLR=TRUE, centerMode=1, scaleMode=2)

# Metabolomics stays the same
processedMetabolomics = parafac4microbiome::vanderPloeg2024$metabolomics

# Homogenize
datasets = list(processedTongue$data, processedLowling$data, processedLowinter$data, processedUpling$data, processedUpinter$data, processedSaliva$data, processedMetabolomics$data)
modes = list(c(1,2,3),c(1,4,5),c(1,6,7),c(1,8,9),c(1,10,11),c(1,12,13),c(1,14,15))
Z = setupCMTFdata(datasets,modes,normalize=TRUE)
```

## Model selection

```{r acmtf model selection}
# Please see the separate R scripts at https://doi.org/10.5281/zenodo.16993009 for the model selection.
```

```{r fit acmtf}
# acmtf_model = CMTFtoolbox::acmtf_opt(Z, 3, nstart = 100, method="L-BFGS", numCores=10)
acmtf_model = readRDS("./TIFN2/acmtf_model.RDS")
acmtf_model$varExp
```

## Description

ACMTF was applied to the capture to dominant source of variation across all data blocks. The model selection procedure revealed that FMS CV dropped below 0.9 when four components were selected. Therefore, a three-component ACMTF model was selected, explaining 26.1%, 13.4%, 7.1%, 10.7%, 12.0%, and 17.1% of the variation in the tongue, lower jaw lingual, lower jaw interproximal, upper jaw lingual, upper jaw interproximal, and salivary microbiome datasets, respectively, and 18.3% of the variation in the salivary metabolomics data.

The Lambda-matrix of the model revealed that the first component was predominantly local to the tongue, lower jaw lingual, upper jaw lingual, and salivary microbiome blocks, while the second component was predominantly local to the salivary metabolomics and tongue microbiome blocks, and the third component was predominantly distinct to the salivary metabolomics block. However, all components received minor contributions from other blocks as well. MLR analysis revealed that the first component captured uninterpretable variation, the second component captured variation associated with a mixture of plaque% and subject age, while the third component captured variation exclusively associated with subject age.

```{r acmtf test the model}
testMetadata(acmtf_model, 1, processedTongue)
testMetadata(acmtf_model, 2, processedTongue)
testMetadata(acmtf_model, 3, processedTongue)
```

```{r acmtf plot the lambda matrix}
lambda = abs(acmtf_model$Fac[[16]])
colnames(lambda) = paste0("C", 1:3)

lambda %>%
  as_tibble() %>%
  mutate(block=c("tongue", "lowling", "lowinter", "upling", "upinter", "saliva", "metab")) %>%
  mutate(block=factor(block, levels=c("tongue", "lowling", "lowinter", "upling", "upinter", "saliva", "metab"))) %>%
  pivot_longer(-block) %>%
  ggplot(aes(x=as.factor(name),y=value,fill=as.factor(block))) +
  geom_bar(stat="identity",position=position_dodge(),col="black") +
  xlab("ACMTF component number") +
  ylab(expression(lambda)) +
  scale_x_discrete(labels=1:3) +
  scale_fill_manual(name="Dataset",values = hue_pal()(7),labels=c("Tongue", "Lower jaw, lingual", "Lower jaw, interproximal", "Upper jaw, lingual", "Upper jaw, interproximal", "Salivary microbiome", "Salivary metabolome")) +
  theme(legend.position="top")
```

```{r acmtf comp 2 feature loadings}
# Component 2 is interpretable as being related to age
# it is mainly described by tongue + metab

temp = list()
temp$Fac = acmtf_model$Fac[c(1,2,3)]
testFeatures(temp, processedTongue, 2, "age")
a = plotFeatures(processedTongue$mode2, temp, 2, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(5,6,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,4,5)]
testFeatures(temp, processedLowling, 2, "age")
b = plotFeatures(processedLowling$mode2, temp, 2, flip=FALSE) + scale_fill_manual(values=phylum_colors[-7])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,6,7)]
testFeatures(temp, processedLowinter, 2, "age")
c = plotFeatures(processedLowinter$mode2, temp, 2, flip=FALSE)

temp = list()
temp$Fac = acmtf_model$Fac[c(1,8,9)]
testFeatures(temp, processedUpling, 2, "age")
d = plotFeatures(processedUpling$mode2, temp, 2, flip=FALSE)

temp = list()
temp$Fac = acmtf_model$Fac[c(1,10,11)]
testFeatures(temp, processedUpinter, 2, "age")
e = plotFeatures(processedUpinter$mode2, temp, 2, flip=FALSE)

temp = list()
temp$Fac = acmtf_model$Fac[c(1,12,13)]
testFeatures(temp, processedSaliva, 2, "age") # flip
f = plotFeatures(processedSaliva$mode2, temp, 2, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,14,15)]
testFeatures(temp, processedMetabolomics, 2, "age")

# Metab
df = processedMetabolomics$mode2 %>% mutate(Component_2 = acmtf_model$Fac[[14]][,2]) %>% arrange(Component_2) %>% mutate(index=1:nrow(.))
df = rbind(df %>% head(n=10), df %>% tail(n=10))

g=df %>%
  ggplot(aes(x=Component_2,y=as.factor(index),fill=as.factor(Type))) +
  geom_bar(stat="identity", col="black") +
  scale_y_discrete(label=df$Name) +
  xlab("Loading") +
  ylab("") +
  guides(fill=guide_legend(title="Class")) +
  theme(text=element_text(size=16), legend.position="none")

g2 = df %>%
  ggplot(aes(x = Component_2, y = as.factor(index),
             fill = as.factor(Type),
             pattern = as.factor(Type))) +
  geom_bar_pattern(stat = "identity",
                   colour = "black",
                   pattern = "stripe",
                   pattern_fill = "black",
                   pattern_density = 0.2,
                   pattern_spacing = 0.05,
                   pattern_angle = 45,
                   pattern_size = 0.2) +
  scale_y_discrete(labels = df$Name) +
  xlab("Loading") +
  ylab("") +
  guides(fill = guide_legend(title = "Class"),
         pattern = "none") +
  theme(text = element_text(size = 16),
        legend.position = "top")

a
b
c
d
e
f
g2
```

```{r acmtf comp 2 time loadings}
a = processedTongue$mode3 %>%
  mutate(Comp = acmtf_model$Fac[[3]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  

b = processedLowling$mode3 %>%
  mutate(Comp = acmtf_model$Fac[[5]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  

c = processedLowinter$mode3 %>%
  mutate(Component_1 = acmtf_model$Fac[[7]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

d = processedUpling$mode3 %>%
  mutate(Component_1 = acmtf_model$Fac[[9]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

e = processedUpinter$mode3 %>%
  mutate(Component_1 = acmtf_model$Fac[[11]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

f = processedSaliva$mode3 %>%
  mutate(Component_1 = -1*acmtf_model$Fac[[13]][,2], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

g = processedMetabolomics$mode3 %>%
  mutate(Component_1 = acmtf_model$Fac[[15]][,2], day=c(0,2,5,9,14)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

a
b
c
d
e
f
g
```

```{r acmtf comp 3 feature loadings}
# Component 2 is interpretable as being related to age
# it is mainly described by tongue + metab

temp = list()
temp$Fac = acmtf_model$Fac[c(1,2,3)]
testFeatures(temp, processedTongue, 3, "age")
a = plotFeatures(processedTongue$mode2, temp, 3, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(3,5,6,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,4,5)]
testFeatures(temp, processedLowling, 3, "age")
b = plotFeatures(processedLowling$mode2, temp, 3, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,6,7)]
testFeatures(temp, processedLowinter, 3, "age") # flip
c = plotFeatures(processedLowinter$mode2, temp, 3, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,8,9)]
testFeatures(temp, processedUpling, 3, "age") # cor says flip, but lambda is negative -> no flip
d = plotFeatures(processedUpling$mode2, temp, 3, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,10,11)]
testFeatures(temp, processedUpinter, 3, "age")
e = plotFeatures(processedUpinter$mode2, temp, 3, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,12,13)]
testFeatures(temp, processedSaliva, 3, "age") # flip
f = plotFeatures(processedSaliva$mode2, temp, 3, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtf_model$Fac[c(1,14,15)]
testFeatures(temp, processedMetabolomics, 3, "age") # flip

# Metab
df = processedMetabolomics$mode2 %>% mutate(Component_2 = -1*acmtf_model$Fac[[14]][,3]) %>% arrange(Component_2) %>% mutate(index=1:nrow(.))
df = rbind(df %>% head(n=10), df %>% tail(n=10))

g=df %>%
  ggplot(aes(x=Component_2,y=as.factor(index),fill=as.factor(Type))) +
  geom_bar(stat="identity", col="black") +
  scale_y_discrete(label=df$Name) +
  xlab("Loading") +
  ylab("") +
  guides(fill=guide_legend(title="Class")) +
  theme(text=element_text(size=16), legend.position="none")

g2 = df %>%
  ggplot(aes(x = Component_2, y = as.factor(index),
             fill = as.factor(Type),
             pattern = as.factor(Type))) +
  geom_bar_pattern(stat = "identity",
                   colour = "black",
                   pattern = "stripe",
                   pattern_fill = "black",
                   pattern_density = 0.2,
                   pattern_spacing = 0.05,
                   pattern_angle = 45,
                   pattern_size = 0.2) +
  scale_y_discrete(labels = df$Name) +
  xlab("Loading") +
  ylab("") +
  guides(fill = guide_legend(title = "Class"),
         pattern = "none") +
  theme(text = element_text(size = 16),
        legend.position = "top")

a
b
c
d
e
f
g2
```

```{r acmtf comp 3 time loadings}
a = processedTongue$mode3 %>%
  mutate(Comp = acmtf_model$Fac[[3]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  

b = processedLowling$mode3 %>%
  mutate(Comp = acmtf_model$Fac[[5]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  

c = processedLowinter$mode3 %>%
  mutate(Component_1 = -1*acmtf_model$Fac[[7]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

d = processedUpling$mode3 %>%
  mutate(Component_1 = -1*acmtf_model$Fac[[9]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

e = processedUpinter$mode3 %>%
  mutate(Component_1 = acmtf_model$Fac[[11]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

f = processedSaliva$mode3 %>%
  mutate(Component_1 = -1*acmtf_model$Fac[[13]][,3], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(-1,1)

g = processedMetabolomics$mode3 %>%
  mutate(Component_1 = -1*acmtf_model$Fac[[15]][,3], day=c(0,2,5,9,14)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

a
b
c
d
e
f
g
```

# ACMTF-R

## Processing

```{r prepare y}
Y = processedTongue$mode1 %>% left_join(rf_data %>% select(subject,Area_delta_R30,day) %>% filter(day==14)) %>% select(Area_delta_R30) %>% pull()
Ycnt = Y - mean(Y)
Ynorm = Ycnt / norm(Ycnt, "2")
```

## Model selection

```{r acmtfr model selection}
# Please see the separate R scripts at https://doi.org/10.5281/zenodo.16993009 for the model selection.
```

```{r acmtfr load models}
acmtfr_model80 = readRDS("./TIFN2/TIFN2_ACMTFR_model_0.8.RDS")
acmtfr_model80$varExp
```

## Description

ACMTF-R was next applied to supervise the joint decomposition using the percentage of sites covered in red fluorescent plaque on day 14 (RF%) as dependent variable (y). The model selection procedure revealed that two components were optimal for pi=0.9, based on high and stable FMS random across datasets, FMS CV values greater than 0.9, and only a marginal increase in RMSECV compared to one component. For pi=0.85, 0.8, and 0.5, one component was chosen, as higher-component models showed sharply declining FMS random and FMS CV scores, suggesting poor model stability. Subsequently, the one-component ACMTF-R model using pi=0.80 was selected for further interpretation due to balancing explaining the independent data and predicting y. This model explained 9.3%, 8.1%, 3.6%, 6.8%, 6.8%, 6.3%, 3.0% and 52.3% of the variation in the tongue, lower jaw lingual, lower jaw interproximal, upper jaw lingual, upper jaw interproximal, salivary microbiome, salivary metabolome, and y, respectively.

The Lambda-matrix of the model revealed that the component was predominantly local to the tongue and lower jaw lingual microbiomes, with moderate contributions from the upper jaw lingual, upper jaw interproximal, and salivary microbiome blocks. As expected, MLR analysis revealed that the component was exclusively associated with RF%.

```{r acmtfr lambda matrices}
lambda = abs(acmtfr_model80$Fac[[16]])
colnames(lambda) = paste0("C", 1)

lambda %>%
  as_tibble() %>%
  mutate(block=c("tongue", "lowling", "lowinter", "upling", "upinter", "saliva", "metab")) %>%
  mutate(block=factor(block, levels=c("tongue", "lowling", "lowinter", "upling", "upinter", "saliva", "metab"))) %>%
  pivot_longer(-block) %>%
  ggplot(aes(x=as.factor(name),y=value,fill=as.factor(block))) +
  geom_bar(stat="identity",position=position_dodge(),col="black") +
  xlab("ACMTF-R component number") +
  ylab(expression(lambda)) +
  scale_x_discrete(labels=1:3) +
  scale_fill_manual(name="Dataset",values = hue_pal()(7),labels=c("Tongue", "Lower jaw, lingual", "Lower jaw, interproximal", "Upper jaw, lingual", "Upper jaw, interproximal", "Salivary microbiome", "Salivary metabolome")) +
  theme(legend.position="top")
```

```{r acmtfr test the model}
testMetadata(acmtfr_model80, 1, processedTongue)
```

```{r acmtfr feature mode}
# Pi = 0.8
temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,2,3)]
testFeatures(temp, processedTongue, 1, "Area_delta_R30") # flip
a = plotFeatures(processedTongue$mode2, temp, 1, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,4,5)]
testFeatures(temp, processedLowling, 1, "Area_delta_R30") # flip
b = plotFeatures(processedLowling$mode2, temp, 1, flip=TRUE) + scale_fill_manual(values=phylum_colors[-7])

temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,6,7)]
testFeatures(temp, processedLowinter, 1, "Area_delta_R30") # flip
c = plotFeatures(processedLowinter$mode2, temp, 1, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,8,9)]
testFeatures(temp, processedUpling, 1, "Area_delta_R30")
d = plotFeatures(processedUpling$mode2, temp, 1, flip=FALSE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,10,11)]
testFeatures(temp, processedUpinter, 1, "Area_delta_R30") # flip
e = plotFeatures(processedUpinter$mode2, temp, 1, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,12,13)]
testFeatures(temp, processedSaliva, 1, "Area_delta_R30") # flip
f = plotFeatures(processedSaliva$mode2, temp, 1, flip=TRUE) + scale_fill_manual(values=phylum_colors[-c(3,7)])

# Metab
temp = list()
temp$Fac = acmtfr_model80$Fac[c(1,14,15)]
testFeatures(temp, processedMetabolomics, 1, "Area_delta_R30") # no flip

df = processedMetabolomics$mode2 %>% mutate(Component_2 = acmtfr_model80$Fac[[14]][,1]) %>% arrange(Component_2) %>% mutate(index=1:nrow(.))
df = rbind(df %>% head(n=10), df %>% tail(n=10))

g = df %>%
  ggplot(aes(x = Component_2, y = as.factor(index),
             fill = as.factor(Type),
             pattern = as.factor(Type))) +
  geom_bar_pattern(stat = "identity",
                   colour = "black",
                   pattern = "stripe",
                   pattern_fill = "black",
                   pattern_density = 0.2,
                   pattern_spacing = 0.05,
                   pattern_angle = 45,
                   pattern_size = 0.2) +
  scale_y_discrete(labels = df$Name) +
  xlab("Loading") +
  ylab("") +
  guides(fill = guide_legend(title = "Class"),
         pattern = "none") +
  theme(text = element_text(size = 16),
        legend.position = "top")

a
b
c
d
e
f
g
```

```{r acmtfr time mode}
a = processedTongue$mode3 %>%
  mutate(Comp = acmtfr_model80$Fac[[3]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

b = processedLowling$mode3 %>%
  mutate(Component_1 = acmtfr_model80$Fac[[5]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

c = processedLowinter$mode3 %>%
  mutate(Component_1 = acmtfr_model80$Fac[[7]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

d = processedUpling$mode3 %>%
  mutate(Component_1 = -1*acmtfr_model80$Fac[[9]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)

e = processedUpinter$mode3 %>%
  mutate(Component_1 = acmtfr_model80$Fac[[11]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  ggplot(aes(x=day,y=Component_1)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  
f = processedSaliva$mode3 %>%
  mutate(Comp = acmtfr_model80$Fac[[13]][,1], day=c(-14,0,2,5,9,14,21)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  
g = processedMetabolomics$mode3 %>%
  mutate(Comp = acmtfr_model80$Fac[[15]][,1], day=c(0,2,5,9,14)) %>%
  select(-visit,-status) %>%
  pivot_longer(-day) %>%
  ggplot(aes(x=day,y=-1*value)) +
  annotate(geom = "rect", xmin = 0, xmax = 14, ymin = -Inf, ymax = Inf, fill = "red", colour = "black", alpha=0.25) +
  geom_line() +
  geom_point() +
  theme(legend.position="none") +
  xlab("Time point [days]") +
  ylab("Loading") +
  ylim(0,1)
  
a
b
c
d
e
f
g
```
Across the microbiome datasets, consistently positively loaded taxa including Aggregatibacter, Fusobacterium, Leptotrichia, Peptostreptococcus, Porphyromonas, and Fusobacterium nucleatum were enriched in high responders and depleted in low responders. Conversely, consistently negatively loaded taxa including Kingella, Lautropia, Prevotella, Rothia, Streptococcus, and Veillonella were enriched in low responders and enriched in high responders. Overall, inter-subject differences due to RF% remained relatively constant, with a small increase during the intervention.

Higher RF% was associated with elevated levels of amino acids, including valine, isoleucine, and methionine. Conversely, individuals with lower RF% exhibited higher levels of peptides (e.g., uridine, inosine, and guanosine) and some carbohydrates (e.g., glucose and maltose). The time mode loadings indicate that inter-subject differences due to the intervention remained relatively constant.
